---
title: "form.io - Nicely formatted objects output in R Markdown"
author: "Philippe Grosjean (phgrosjean@sciviews.org)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{form.io - Nicely formatted objects output in R Markdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Among the 'SciViews' R packages, 'form.io' would be devoted to the input and especially output of textual data (input of tabular data being managed by 'data.io'). Here, we reuse existing functions like `knitr::kable()` or `pander::pander()`, but we make them simpler to use and working in all cases being:

- At the R console,
- Inside an R Markdown/Notebook inline area just beneath a chunk,
- In a knitted R markdown file, HTML format,
- Idem, but with a LaTeX or Word format.

Both 'knitr' with `kable()`, and the 'kableExtra' functions, and 'pander' with the `pander()` functions do that. However, the syntax is not always easy, and code eventually differ depending on the situation among the four possible ones here above.

Moreover, none of `kable()` or `pander()` functions allow for the *translation* of the result in a different language, and they also do not take into account attributes like `label` or `units` (see the 'data.io' package).

So, the new generic function `form()` creates formatted tables or other objects (lists, ...) to be directly integrated in an R Markdown document, or to be used at the R Console. Just load the package (and perhaps also 'data.io', or even the whole `SciViews::R` suite):

```{r}
# Better to use SciViews::R() here
library(data.io)
library(form.io)
# To change language to, e.g., French
options(data.io_lang = "FR") # Same language option is used by form.io::form()
```

```{r}
# Test
trees <- read("trees", package = "datasets")
lm_1 <- lm(data = trees, volume ~ .)
form(trees)
form(head(trees), data = trees) # Try without data =
form(anova(lm_1), data = trees)
```

### Using `form()`

`pander::pander()` works well with knitted documents, but it outputs text only in R notebooks. The `form()` function is solving this problem, while still outputting a textual markdown version at the R console.

```{r}
trees <- read("trees", package = "datasets")

lm_1 <- lm(data = trees, volume ~ .)
lm_2 <- lm( data = trees, volume ~ I(diameter^2))
```

- Data frame\ :

```{r}
form(trees, caption = "Mesure de cerisiers noirs")
# Equivalent to:
#form(trees, caption = "Mesure de cerisiers noirs", labels = TRUE, units = TRUE)
```

Any manipulation of a data frame with labels and units makes it loose its attributes, ... but the `data =` argument is there to look at original data set for these attributes\ :

```{r}
form(head(trees))
form(head(trees), data = trees)
```

- Objects `data.table`\ :

```{r}
library(data.table)
trees_dt <- as.data.table(trees)
class(trees_dt)
form(trees_dt)
```

- Object `lm`\ :

```{r}
form(lm_1, data = trees, signif_stars = TRUE)
```

- Summary of `lm`\ :

```{r}
form(summary(lm_1))
```

- Anova of `lm`\ :

```{r}
form(anova(lm_1), data = trees) #, caption = "ANOVA de ma regression linéaire")
```

- Generalized linear models\ :

```{r}
# From ?glm
## Dobson (1990) Page 93: Randomized Controlled Trial :
ad <- data.frame(treatment = gl(3, 3), outcome = gl(3, 1, 9),
  counts = c(18, 17, 15, 20, 10, 20, 25, 13, 12))
glm_1 <- glm(data = ad, counts ~ outcome + treatment, family = poisson())
glm_1
summary(glm_1)
anova(glm_1)
```

```{r}
anorexia <- read("anorexia", package = "MASS")
form(anorexia)
glm_2 <- glm(data = anorexia, Postwt ~ Prewt + Treat, family = gaussian)
glm_3 <- glm(data = anorexia, Postwt ~ Prewt + Treat + offset(Prewt), family = gaussian)
form(glm_2)
form(summary(glm_2))
#form(anova(glm_2))
form(aov(glm_2))
```

```{r}
# A Gamma example, from McCullagh & Nelder (1989, pp. 300-2)
clotting <- data.frame(
  u = c(5, 10, 15, 20, 30, 40, 60, 80, 100),
  lot1 = c(118, 58, 42, 35, 27, 25, 21, 19, 18),
  lot2 = c(69, 35, 26, 21, 18, 16, 13, 12, 12))
glm_4 <- glm(data = clotting, lot1 ~ log(u), family = Gamma)
glm_4
summary(glm_4)
anova(glm_4)

glm_5 <- glm(data = clotting, lot2 ~ log(u), family = Gamma)
glm_5
summary(glm_5)
anova(glm_5)

## Aliased ("S"ingular) -> 1 NA coefficient
(fS <- glm(lot2 ~ log(u) + log(u^2), data = clotting, family = Gamma))
tools::assertError(update(fS, singular.ok = FALSE), verbose = interactive())
## -> .. "singular fit encountered"
```

- Nonlinear regression objects\ :

```{r}
nls_1 <- nls(data = trees, volume ~ SSlogis(diameter, Asymptote, Inflexion, Echelle))
form(nls_1)
form(nls_1, params_labels = c("Asym", "xmoyen", "scal")) # Rename the parameters
form(summary(nls_1))
```

```{r}
nls_2 <- nls(data = trees, volume ~ SSgompertz(diameter, Asymptote, B2, B3))
form(anova(nls_1, nls_2))
```

- Lists\ :

```{r}
form(list(a = 1:3, b = TRUE, c = LETTERS))
```

- Objects `ts`\ :

```{r}
form(nottem, caption = "températures à Notthingam")
```

- Unknown object type (coerced to a list with a warning)\ :

```{r}
form(structure(list(a = 1, b = 3:5), class = "new_oject"))
```

### Using `kable()`

Note: the `kable()` functions are not integrated yet into `form()` for the moment! The `kable()` function is standard in 'knitr'. Moreover, the 'kableExtra' package allows for further configuration of the output.

```{r}
#options( knitr.table.format = "markdown")
knitr::kable(trees, caption = "The `trees` dataset")
#kableExtra::xtable2kable(xtable::xtable(linear.1))
```

```{r}
knitr::kable(broom::glance(lm_1)[, c(1, 3, 5, 8, 11)])
```

```{r}
knitr::kable(broom::tidy(lm_1), caption = "Tableau des coefficients")
```

### Using `stargazer`

Not good! But we may considering to fix these...

```{r}
#stargazer::stargazer(trees, type = "text")
#stargazer::stargazer(trees, type = "latex")
#stargazer::stargazer(lm_1, type = "text")
```
